Tracking Rodinia into the Neoproterozoic: new paleomagnetic constraints from the Jacobsville Formation

### 2023-03-29 Manuscript/Manuscript.tex
update manuscript tex files

\begin{document}

\title{Laurentia in the earliest Neoproterozoic: new paleomagnetic constraint from the Jacobsville Formation}

\authors{Yiming Zhang\textsuperscript{1}, Blake Hodgin\textsuperscript{1}, Nicholas Swanson-Hysell\textsuperscript{1}, James Pierce\textsuperscript{1,2}, Anthony Fuentes\textsuperscript{1}}

\affiliation{1}{Department of Earth and Planetary Science, University of California, Berkeley, CA, USA}
\affiliation{2}{Department of Earth and Planetary Sciences, Yale University, CT, USA}

@@

\begin{keypoints}
\item A new paleomagnetic pole from the ca. 990 Ma Jacobsville Formation \item Laurentia crossed the equator in the latest Mesoproterozoic to earliest Neoproterozoic
\item The new pole indicates the Grenville Loop could be much younger than thought
\end{keypoints}

\begin{abstract}
The paleogeography of Laurentia in the late Proterozoic is central to global paleogeography due to its central position in the supercontinent Rodinia. We develop a new paleomagnetic pole from red beds of the ca. 990 Ma Jacobsville Formation which serves as a crucial anchor point for Laurentia's position in the earliest Neoproterozoic. High-resolution thermal demagnetization experiments resolve detrital remanent magnetizations that pass a conglomerate test. An inclination corrected paleomagnetic mean pole position from the Jacobsville Formation confirm continued plate motion of Laurentia across the equator in the late Mesoproterozoic to early Neoproterozoic. The new pole from the interior of Laurentia indicates the plate slowed down significantly following the onset of the Grenvillian orogenesis on its eastern margin. This pole's position and age indicate the paleomagnetic poles developed from the Grenville Province most likely record coherent plate motion after major continental collision during the initial assembly of the supercontinent Rodinia.
\end{abstract}

% \section*{Plain Language Summary}
% [ enter your Plain Language Summary here or delete this section]

\section*{Keywords}
Rodinia\\
Laurentia  \\
Paleogeography\\
Neoproterozoic\\
\

apparent polar wander path\\
Inclination shallowing\\
Grenville\\
Keweenawan superchron\\
Red bed\\

@@

The extensively studied paleomagnetic records of the ca. 1109 Ma to 1084 Ma North American Midcontinent Rift (i.e. the Keweenawan Track) provide crucial constraints on the paleogeography of Laurentia in the Mesoproterozoic \cite{Swanson-Hysell2019a}. As a result, the Keweenawan Track has become a central record for reconstructing the initial development of the ancient supercontinent Rodinia \cite{2021a}. As the rifting ceased and subsequently inverted following the onset of the Grenvillian Orogeny on Laurentia's eastern margin (Fig. \ref{fig:Geologic_map}), magmatism in the interior of Laurentia entered a period of quiescence that could have lasted for as long as $\sim$300 Myr until the emplacement of the ca. 780 Ma Gunbarrel large igneous province \cite{Harlan2003a, Mackinder2019a}. Due to this magmatic gap, Laurentia and its conjugate continents' paleogeographic configuration in the late Mesoproterozoic to mid-Neoproterozoic has remained largely uncertain \cite{Swanson-Hysell2021c}.

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{Geologic_map.pdf}
\caption{(A) Regional map showing location of the Grenville Front relative to the study area. The inset green box shows the extent of panel B. (B) Geologic map of northern Michigan and Wisconsin showing Midcontinent Rift igneous and sedimentary rocks, the Jacobsville Formation, other older Proterozoic rocks, and major post-rift thrust faults. Jacobsville stratigraphic sections along which paleomagnetic samples were collected in this study are shown as green diamond symbols.}
\label{fig:Geologic_map}
\end{figure*}

When igneous rock records are lacking, paleomagnetic data recorded by sedimentary rocks (e.g. red beds) can provide useful insights into the ancient geomagnetic field and record paleogeographic positions of sedimentary basins. In the Midcontinent Rift, the paleomagnetic poles developed from the ca. 1075 Ma Nonesuch Formation and the ca. 1070 Ma upper Freda Formation of the Oronto Group sediments that overlie the Midcontinent Rift volcanics in southern Lake Superior area (Fig. \ref{fig:Geologic_map}) provided crucial constraints for Laurentia's paleolatitude, orientation, and plate speed in the late Mesoproterozoic \cite{Henry1977a, Swanson-Hysell2019a, Rose2022a}.

On top of the Oronto Group sediments overlies the post-rift Jacobsville Formation which was partly deformed during the inversion of the Midcontinent Rift (Fig. \ref{fig:Geologic_map}; \citeA{Hamblin1958a, Hodgin2022a, DeGraff2022a}). U-Pb detrital zircon ages and U-Pb calcite ages from the Keweenaw fault where the Jacobsville Formation is deformed against thrusted rift volcanics constrain the maximum depositional age of the Jacobsville Formation to be ca. 990 Ma \cite{Hodgin2022a}. The Jacobsville Formation contains abundant hematite-rich red beds which bear potential to further extend the apparent polar wander path (APWP) of Laurentia toward more recent times \cite{Hamblin1958a, Roy1978a}. Developing paleomagnetic data from the Jacobsville Formation will allow further development of Rodinia's paleogeographic evolution in the late Mesoproterozoic to early Neoproterozoic.

Challenges exist in interpreting sedimentary paleomagnetic records. Primary detrital remanence acquired through settling and deposition of magnetic particles in the water column can be masked by secondary remanence acquired through precipitation and growth of new minerals within pore spaces of host sedimentary rocks syn- to post-deposition (e.g. pigmentary hematite). In red beds, it has been found that formation of the pigmentary hematite can significantly post-date the timing of sediment deposition, resulting in a chemical remanent magnetization (CRM) superimposed upon primary detrital remanent magnetization (DRM) carried by detrital grains \cite{Butler1992a}. To isolate the remanence components of the red beds of the Jacobsville Formation, \citeA{Roy1974a, Roy1978a} adopted a method first applied by \citeA{Collinson1965c} to preferentially remove fine-grained pigmentary hematite through prolonged immersion in concentrated HCl acid whereas leaving the coarser detrital grains that are more resistant to dissolution in place. Studies applying paired acid etching demagnetization and thermal demagnetization show that the pigmentary hematite grains tend to have lower unblocking temperatures than detrital ones \cite{Tauxe1980a, Bilardello2010c}. High-resolution thermal demagnetization experiments with temperature intervals as small as 1-2 \textdegree C have shown to be effective in isolating detrital remanence from chemical remanence as coarser hematite grains tend to have higher unblocking temperatures closer to the Ne\'el temperature of hematite \cite{Swanson-Hysell2019b}. In particular, a conglomerate test on a suite of fluvial intraclasts of the red beds in the Freda Formation shows that via thermal demagnetization one can isolate detrital magnetization vectors (which are reoriented in random directions in the clasts) from the secondary components carried by co-occurring authigenic hematite (which yield similar directions amongst clasts) \cite{Swanson-Hysell2019b}.

Paleomagnetism of red beds has also been challenged by the controversial issue of correcting for inclination shallowing \cite{King1955a, Tauxe2004b, Bilardello2016b}. Due to rotation of detrital magnetic grains caused by authigenic processes such as compaction during lithification, inclinations recorded by sediments often have shallower angles than the local field in which they are deposited. To estimate the amount of inclination shallowing in a given package of sedimentary rocks, \citeA{Tauxe2004b} developed a statistical model (i.e. the elongation/inclination [E/I]method) that evaluates the deviation of the distribution of a large number ($>$100) of observed sedimentary paleomagnetic directional data with the predicted distribution given by a paleosecular variation model developed for the past 5 Myr (TK03) under the the geocentric axial dipole (GAD) hypothesis. Although the TK03 paleosecular variation model is based on data of relatively recent geomagnetic field variations, it has been shown to be compatible with data from the ca. 1.1 Ga lava flows of the Midcontinent Rift \cite{Tauxe2009a}. Further, \citeA{Pierce2022a} showed that the E/I method passes a field test for the 1.1 Ga Cut Face Creek Sandstone in the Midcontinent Rift as it successfully corrects the shallow paleomagnetic inclinations of the sediments to that of the North Shore Volcanic Group lava flows that bracket them. That study also further extended the method to better represent uncertainties associated with the amount of inclination shallowing when reporting sedimentary paleomagnetic pole positions by using the spherical elliptical Kent distribution \cite{Kent1982a} instead of the circular Fisher distribution \cite{Fisher1953a}.

In this study, we investigate the hematite-bearing, fine-grained sandstone to siltstone lithofacies along five stratigraphic sections of the Jacobsville Formation on Keweenaw Peninsula, Michigan, USA. High-resolution thermal demagnetization isolate primary DRM from secondary CRM with the DRM passing a conglomerate test. We then use a paleomagnetic fold test to further constrain the timing of DRM acquisition in association with the structural inversion of the Midcontinent Rift. Finally, we develop a new inclination-corrected paleomagnetic pole for the Jacobsville Formation based on a large number of samples and investigate the paleo-continental configuration of Laurentia in the late Mesoproterozoic to early Neoproterozoic.

% The end of the ca. 1109 Ma to 1084 Ma intracontinental Midcontinent Rift magmatism and extension was marked by a transition to deposition of sedimentary rocks of the Oronto Group \cite{Ojakangas2001a} as the rift basin thermally subsided prior to the rift undergoing contractional deformation in the later stages of the Grenvillian orogeny \cite{Cannon1990b, Hodgin2022a}.

%  that has become central to the reconstruction of paleogeography of the assembly of supercontinent Rodinia. In addition to syn-rift rocks preservation, the cessation and later inversion of the rift allowed the development of syn-to post-rift basin development which accommodated the deposition of the ca. 1080 - 1050 Ma Oronto Group sedimentary rocks that are in total about 5 km thick. Overlying the Oronto Group is the Jacobsville formation which has been interpreted to be deposited in a back-bulge setting associated with the far-field propagation of the crustal-scale deformation caused by the Grenvillian Orogeny.

% A leading tectonic model for the cessation of the Midcontinent Rift and extension is that it occurred due to far-field compressional stress associated with the ca. 1090-980 Ma Grenvillian orogeny \cite{Cannon1990a}, when the collision of Laurentia with its conjugate margins led to the development of a back-bulge basin, accommodating the deposition of the Jacobsville Formation sediments as the inversion of the Midcontinent Rift progressed. Recent re-examination of the maximum deposition age of the Jacobsville Formation using zircon U-Pb whole grain chemical abrasion isotope dilution thermal ionization mass spectrometry has updated the maximum deposition age of the Jacobsville Formation to be ca. 990 Ma \cite{Hodgin2022a}. Such age constraint is consistent with backbulge subsidence model \cite{DeCelles, 2012}, indicating that the Jacobsville Formation may have been deposited in a Grenvillian foreland basin system that resulted from lithospheric flexure induced by orogenic loading \cite{Rivers2012a}. The relatively young age of the Jacobsville Formation compared to other Midcontinent Rift rocks makes the Jacobsville Formation an crucial target for further reconstructing the global paleogeography in the late Proterozoic. For further refining the paleogeographic connection between the poles of the Grenville Loop and those of the Keweenawan Track.

% Paleomagnetic data for the Jacobsville Formation have led to a paleomagnetic pole position that appears to be a continuation of the Keweenawan Track from the Oronto Group poles (Roy and Robertson, 1978; Fig. 7; Table 4). As a result, its age has been interpolated to be latest Mesoproterozoic (e.g., 1020 Ma in Li et al., 2008) such that it was deposited prior to the ca. 1000 Ma Grenville loop poles. However, this age assignment has recently been questioned on the basis of laser-ablation–inductively cou- pled plasma–mass spectrometry (LA-ICP-MS) U-Pb detrital zircon dates from Jacobsville Sandstone samples. Malone et al. (2016) interpreted a maximum depositional age of 959 ± 19 Ma based on the weighted mean of the four youngest grains from 2050 LA-ICP-MS zircon dates. A similar age was proposed on the basis of the youngest LA-ICP-MS U-Pb dates in the study of Craddock et al. (2013). This young maximum interpreted age is intriguing because it suggests the unconformity above the Freda Sandstone spans at least 100 m.y. and was followed by continued sedimentation in the region long after rift activity. Nevertheless, the paleo- magnetic pole of the Jacobsville Sandstone (Roy and Robertson, 1978) appears to extend the Keweenawan Track from the Oronto Group poles in a manner that may be more consistent with deposition of the Jacobsville Formation in the late Mesoproterozoic. Given the conflict with this inferred age, the large uncertainty of individual LA-ICP-MS dates, and the possibility of outlier dates in such a large data set, it is beneficial to interpret the age of the youngest Jacobsville zircons dated CA-ID-TIMS to obtain more precise constraints. Using CA-ID-TIMS, a recent study by Hodgin et al (2022) redefines the maximum deposition age of the Jacobsville Formation as ca. 993 Ma.

% As the Midcontinent Rift magmatic activity wanes its strength and evolved into a failed intracontinental rift, Laurentia experienced a long period of magmatic quiescence in its interior, where basin subsidence dominated the rift basin. This was followed by the Grenville Orogeny, causing the the inversion of the rift along thrust belts such as the Keweenaw Fault (Fig. xxx). The lack of extensive magmatism from ca.1080 Ma to ca. 980 Ma in the Laurentia interior resulted in a significant gap (for at least 50 myr) of paleomagnetic poles between the Keweenawan Track (developed from interior Laurentia igneous and sedimentary rocks) and the Grenville Loop (poles developed from the Grenville Province rocks) (Fig. xxx; \cite{Swanson-Hysell2019a}).

% Due to the lack of pmag records from the interior Laurentia, various hypotheses for the spatial relationship between the end of the Keweenawan Track and the Grenville Loop have been debated. One model argues for that the difference in pole locations indicates that the Grenville Province was once thousands of kilometers away from Interior Laurentia and was progressively approaching the southern margin of Laurentia in the late Mesoproterozoic \citep{Halls2015}. A corollary of this hypothesis is that there had to be a 4000 km shortening during the Grenville orogeny - colliding Amazonia-Laurentia. The other model favors that the Grenville Loop is a continuation of the Keweenawan Track - that the southerly continental drift of Laurentia continued across the equator onto the southern hemisphere from ca. 1070 Ma to ca.1000 Ma. and  approached this problem by developing data from metamorphosed dike and anorthosite and argued for Grenville

\section*{Geologic settings}

\subsection*{Structure}
The ca. 1109 Ma to 1084 Ma North American Midcontinent Rift is a major intracontinental rift system that extends over 2000 km through the Laurentia craton (Fig. \ref{fig:Geologic_map}). However, the rifting failed to separate Laurentia into two continents. As its magmatism waned and thermal subsidence progressed in the rift basin, the $>$4 km of ca. 1085-1050 Ma Oronto Group sediments \cite{Fairchild2017a, Swanson-Hysell2019a} were deposited conformably on top of the rift volcanics (Fig. \ref{fig:Geologic_map}; \citeA{Cannon1989a}). Subsequently, the rift was inverted as the ca. 1090-980 Ma Grenvillian Orogeny commenced on the eastern margin of Laurentia (Fig. \ref{fig:Geologic_map}), eventually resulting in the propagation of far-field compressional stresses deep into the interior of Laurentia \cite{Cannon1993a, Cannon1994a}.

During the rift inversion, Paleoproterozoic and Archean crust was uplifted via thrust faults such as the Marenisco fault and the Pelton Creek fault in the south, forming the crustal-scale Montreal River monocline (Fig. \ref{fig:Geologic_map}; \citeA{Cannon1993a}). The southern part of the post-rift Jacobsville Formation overlies an erosional angular unconformity that developed on lithologies that were exhumed through this earlier reverse motion (Fig. \ref{fig:Geologic_map}; \citeA{Hamblin1958a, Cannon1993a, Kalliokoski1982a}). North to these faults, Midcontinent Rift volcanic rocks were thrusted and folded along the Keweenaw fault which extends $\sim$250km from northern Wisconsin to northern Keweenaw Peninsula (Fig. \ref{fig:Geologic_map}; \citeA{DeGraff2022a}). Final inversion of the Keweenaw fault during the Rigolet phase of the Grenvillian Orogeny deformed part of the Jacobsville strata near the fault \cite{Hodgin2022a}. Recent mapping of the Jacobsville Formation also found that it can sometimes onlap the Midcontinent Rift volcanics within or adjacent to the Keweenaw Fault system near the northern end of the Keweenaw Peninsula (Fig. \ref{fig:Geologic_map}; \citeA{Tyrrell2019a, Mueller2021a}).

>>>>

\begin{document}
\title{Tracking Rodinia into the Neoproterozoic: new paleomagnetic constraints from the Jacobsville Formation}

\authors{Yiming Zhang\textsuperscript{1}, Eben B. Hodgin\textsuperscript{1,2}, Tadesse Alemu\textsuperscript{1}, James Pierce\textsuperscript{1,3}, Anthony Fuentes\textsuperscript{1}, Nicholas L. Swanson-Hysell\textsuperscript{1}}

\affiliation{1}{Department of Earth and Planetary Science, University of California, Berkeley, CA, USA}
\affiliation{2}{Department of Earth, Environmental, and Planetary Science, Brown University, Providence, RI, USA}
\affiliation{3}{Department of Earth and Planetary Sciences, Yale University, CT, USA}

@@

\begin{keypoints}
\item Siltstones and fine-grained sandstones of the ca. 990 Ma Jacobsville Formation have primary magnetization held by detrital hematite.
\item Following closure of the Unimos Ocean and onset of the Grenvillian orogeny in the late Mesoproterozoic, Laurentia's motion significantly slowed.
\item The pole position and age suggests that the Grenville Loop paleomagnetic poles are younger than current interpretations.
\end{keypoints}

\begin{abstract}
The paleogeography of Laurentia throughout the Neoproterozoic is critical for reconstructing global paleogeography due to its central position in the supercontinent Rodinia. We develop a new paleomagnetic pole from red siltstones and fine-grained sandstones of the early Neoproterozoic Jacobsville Formation which is now constrained to be ca. 990 Ma in age. High-resolution thermal demagnetization experiments resolve detrital remanent magnetizations held by hematite. These directions were reoriented within siltstone rip-up clasts and pass an intraformational conglomerate test---giving confidence that the magnetization is detrital and primary. An inclination-corrected mean paleomagnetic pole position for the Jacobsville Formation indicates that Laurentia's motion slowed down significantly following the onset of the Grenvillian orogeny. Prior rapid plate motion associated with closure of the Unimos Ocean transitioned to slow drift of Laurentia across the equator in the late Mesoproterozoic to early Neoproterozoic. We interpret the distinct position of this well-dated pole from those in the Grenville orogen that have been assigned a similar age to indicate that the ages of the poles associated with the Grenville Loop likely need to be revised to be younger due to prolonged exhumation. 
\end{abstract}

\section*{Plain Language Summary}

There have been times in Earth history when many of the continents are joined together into a big continent that is called a supercontinent. The most recent supercontinent is Pangea which had Africa at its center. An older supercontinent called Rodinia formed about 1 billion years ago and had North America at its center. The old parts of North America that were in the supercontinent are called Laurentia. The position of Laurentia is key to understanding where Rodinia was and how different continents were connected in the supercontinent. We now know the age of sedimentary rocks called the Jacobsville Formation that were deposited in ancient rivers and can use these rocks to determine the past position of Laurentia (and therefore Rodinia) 990 million years ago using paleomagnetism. This position is quite different in current interpretations which rely on data from rocks that cooled slowly within an ancient mountain belt that formed when the supercontinent assembled---they likely show where the supercontinent was after 990 Ma. As a result, we have a new understanding of where the supercontinent was that can enable further progress in putting all the pieces together to further understand the ancient Earth.

@@

The extensively studied paleomagnetic records of ca. 1109 to 1084 Ma volcanics and intrusions associated with the North American Midcontinent Rift provide crucial constraints on the paleogeography of Laurentia in the late Mesoproterozoic Era \cite{Halls1982a, Fairchild2017a, Swanson-Hysell2019a}. The resulting sequence of poles known as the Keweenawan Track is a central record for reconstructing the assembly of the ancient supercontinent Rodinia \cite{Evans2021b}. An advantage of intracratonic magmatic events, such as those preserved in the Midcontinent Rift, is that they lead to emplacement of igneous rocks that can readily retain primary thermal remanent magnetization (TRM) that can be confidently associated with the emplacement age of the unit. However, following the development of the Midcontinent Rift, there was a $\sim$300 Myr quiescence in known intracratonic magmatic activity in Laurentia that lasted until the emplacement of the ca. 775 Ma Gunbarrel large igneous province \cite{Harlan2003a, Mackinder2019a, Swanson-Hysell2021c}. The lack of Laurentian rocks with primary TRM in the early Neoproterozoic has led paleogeography reconstructions to be reliant on magnetizations acquired by rocks that were metamorphosed during the ca. 1090 to 980 Ma Grenvillian orogeny \cite{Rivers2012a, Swanson-Hysell2023a}. However, rocks within the orogen acquired their magnetizations during exhumation long after peak metamorphic conditions \cite{McWilliams1975a}. As a result, the timing of remanence acquisition needs to be constrained through the more difficult task of dating cooling associated with exhumation. This gap in well-dated paleomagnetic poles for Laurentia in the late Mesoproterozoic to mid-Neoproterozoic contributes to uncertainty in paleogeography at that time including the configuration of the supercontinent Rodinia.

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{Geologic_map.pdf}
\caption{(A) Regional map showing the extent of the Midcontinent Rift and the location of the Grenville Front relative to the study area. The inset green box shows the extent of panel B. (B) Geologic map of northern Michigan and Wisconsin showing Midcontinent Rift igneous and sedimentary rocks, the Jacobsville Formation, sedimentary rocks of the Bayfield Group, other older Proterozoic rocks, and major post-rift thrust faults. The location of the Jacobsville stratigraphic sections in this study are shown with green diamonds. Map modified from \citeA{Nicholson2004a}.}
\label{fig:Geologic_map}
\end{figure*}

Paleomagnetic poles can also be developed from sedimentary rocks deposited in basins. Sedimentary rocks of the Oronto Group deposited during post-rift thermal subsidence following the bulk of Midcontinent Rift magmatic activity provide paleomagnetic poles that extend the Keweenawan Track to ca. 1070 Ma \cite{Henry1977a, Slotznick2023a}. While there are some contemporaneous basins in northern Laurentia \cite{Greenman2021a}, there is minimal basin development within Laurentia following the Grenvillian orogeny until the Rodinia supercontinent began to rift apart ca. 770 Ma \cite{Macdonald2023a}.

One rare preservation of an early Neoproterozoic sedimentary succession in Laurentia's interior is the Jacobsville Formation (Fig. \ref{fig:Geologic_map}; \citeA{Hamblin1958a, Hodgin2022a, DeGraff2022a}). A 992.51$\pm$0.64 Ma (2$\sigma$ analytical uncertainty) U-Pb detrital zircon date from the Jacobsville Formation constrains its maximum deposition age \cite{Hodgin2022a}. A 985.5$\pm$35.8 Ma (2$\sigma$ analytical uncertainty) U-Pb date from calcite veins within the Keweenaw fault provides a minimum age constraint on Jacobsville deposition, as the Jacobsville Formation is folded in the footwall against Midcontinent rift volcanics in the hanging wall. Together, these dates bracket the depositional age of the Jacobsville Formation and constrain it to be ca. 990 Ma \cite{Hodgin2022a}. These results constrain the formation to have been deposited within a syn-orogenic basin during the Rigolet phase of the Grenvillian orogen and to have been deformed during the same orogenic phase \cite{Hodgin2022a, Swanson-Hysell2023a}.

The Jacobsville Formation contains abundant hematite-bearing sandstone and siltstone which have the potential to record a key pole in Laurentia's apparent polar wander path (APWP) \cite{Dubois1962a, Roy1978a}. \citeA{Roy1978a} conducted a paleomagnetic study that developed a pole which has been used in paleogeographic syntheses where it has been assigned variable ages including an age of ca. 1020 Ma based on APWP extrapolation \cite<e.g.>{Li2008a}. However, the uncertainty in its age combined with a lack of field tests, and no application of inclination correction have resulted in the pole not being included in recent curated compilations \cite<e.g.>{Evans2021a}. Obtaining high-quality paleomagnetic data, including robust field tests, motivates the development of new data from the Jacobsville Formation.

In this study, we investigate five stratigraphic sections of the Jacobsville Formation on and near the Keweenaw Peninsula, Michigan, USA (Figs. \ref{fig:Geologic_map} and \ref{fig:strat_column}). High-resolution thermal demagnetization isolates primary detrital remanent magnetization (DRM) from secondary chemical remanent magnetization (CRM) within analyzed samples. This interpretation is supported by the DRM directions passing an intraformational conglomerate test on siltstone intraclasts that were ripped-up and remobilized within the fluvial depositional environment. A paleomagnetic fold test further constrains the timing of DRM acquisition to predate the folding of the Jacobsville Formation in the footwall of the Keweenaw Fault soon after deposition. With these data, we develop a new inclination-corrected paleomagnetic pole for the Jacobsville Formation based on a large number of samples that gives new constraints on the paleogeographic position of Laurentia in the earliest Neoproterozoic.

\section*{Geologic setting}

The ca. 1109 Ma to 1084 Ma North American Midcontinent Rift is a major intracontinental rift system that extends over 2000 km through the Laurentia craton (Fig. \ref{fig:Geologic_map}). Following the end of active extension in the rift, thermal subsidence resulted in the accumulation of $>$4 km of ca. 1085-1050 Ma Oronto Group sedimentary rocks that are both intercalated with the youngest volcanics and deposited atop them (Fig. \ref{fig:Geologic_map}; \citeA{Daniels1982a, Cannon1989a}). Subsequently, the rift was inverted as contraction associated with the ca. 1090-980 Ma Grenvillian Orogeny on the eastern margin of Laurentia (Fig. \ref{fig:Geologic_map}) propagated into Laurentia's interior \cite{Cannon1993a, Cannon1994a, Hodgin2022a, Swanson-Hysell2023a}.

During rift inversion, Midcontinent Rift volcanic and sedimentary rocks were uplifted along with Paleoproterozoic and Archean lithologies via thrust faults such as the Marenisco fault, forming the crustal-scale Montreal River monocline (Fig. \ref{fig:Geologic_map}; \citeA{Cannon1993a}). Rb-Sr radiometric thermochronology data of \citeA{Cannon1993a} show Archean lithologies within the Montreal River monocline to have been exhumed to mid to shallow crustal temperatures ($\sim$270\textdegree C) by ca. 1050 Ma. The Jacobsville Formation overlies an erosional angular unconformity that developed on lithologies that were exhumed through this earlier episode of contractional deformation associated with Grenvillian orogenesis (Fig. \ref{fig:Geologic_map}; \citeA{Hamblin1958a, Cannon1993a, Kalliokoski1982a}). Conglomeratic facies can occur in the basal Jacobsville Formation, with clasts derived from locally uplifted basement lithologies \cite{Irving1885a, Hamblin1958a, Kalliokoski1982a, Alemu2023a}.

Bedding planes in the Jacobsville Formation typically have shallow dips. The exception to these near-horizontal orientations is proximal to the Keweenaw fault (Fig. \ref{fig:Geologic_map}) where the Jacobsville Formation was occasionally deformed into monoclinal drag folds resulting in steeply tilted to overturned beds (Fig. \ref{fig:Geologic_map}; \citeA{Irving1885a, Cannon2001a}). Recent mapping of the Jacobsville Formation also found that it sometimes onlaps the Midcontinent Rift volcanics within or adjacent to en echelon fault branches of the Keweenaw fault system near the northern end of the Keweenaw Peninsula \cite{Tyrrell2019a, Mueller2021a}. Clumped isotope data for vein calcite with a U-Pb date of 985.5$\pm$35.8 Ma in the Keweenaw fault breccia within the hanging wall give temperatures of $\sim$50\textdegree C  \cite{Hodgin2022a}. The low temperature indicates that the calcite precipitated at shallow depth and that the majority of contractional inversion along the fault was complete by ca. 985 Ma. Combined with the 992.51$\pm$0.64 Ma maximum depositional age, these data are interpreted to indicate that the Jacobsville is a syn-orogenic unit deposited during the ca. 1010-980 Ma Rigolet phase of the Grenvillian orogeny \cite{Hodgin2022a, Swanson-Hysell2023a}.

In addition, field observations reveal erosional contacts and development of paleosols between the Jacobsville Formation and underlying volcanic rocks \cite{Hamblin1958a, Kalliokoski1975a}, and seismic reflection data beneath Lake Superior have been interpreted to indicate that the Jacobsville Formation lies in angular unconformity atop sedimentary rocks of the Oronto Group \cite{Cannon1989a}. These data indicate that there was an interval of erosion following ca. 1050 Ma prior to the onset of Jacobsville Formation deposition.

\subsection*{Overview of Jacobsville sedimentology and stratigraphy}

The Jacobsville Formation is a fluvial sequence of feldspathic and quartzose conglomerates, sandstones, siltstones, and shales devoid of lava flows or cross-cutting igneous dikes although internal clastic dikes are present \cite{Hamblin1958a}. Sandstones of the Jacobsville Formation are more quartz-rich with fewer lithic fragments than sandstones of the Oronto Group such that they can be classified as subarkose to quartz arenites (Fig. \ref{fig:Field_photo}; \citeA{Hamblin1958a, Ojakangas2002a}). While much of the Jacobsville Formation is dominated by sandstone channel deposits, there are clay-rich siltstones that formed in overbank settings that are a major focus of this study.

\begin{figure*}[h!]
\centering
\includegraphics[width=0.8\textwidth]{Jacobsville_Sections_v6.pdf}
\caption{Lithostratigraphy and magnetostratigraphy for studied sections of the Jacobsville Formation in northern Michigan, USA. Paleomagnetic sampling focused on the dark red siltstone to fine-grained sandstone lithofacies. Paleomagnetic specimen declinations are plotted with blue circles corresponding to normal polarity and orange circles to reverse polarity. The maximum depositional age of the Agate Falls section is constrained by a CA-ID-TIMS detrital zircon U-Pb date of 1003.2 $\pm$ 2.2 Ma (sample AF1-29.3 of \citeA{Hodgin2022a}). In the Snake Creek tributary and Natural Wall sections, the position of the detrital zircon samples BSC1-92.5 and NW1–61.5 from \citeA{Hodgin2022a} are shown. No U-Pb zircon dates from these samples were younger than the age of Midcontinent Rift volcanics. A correlation of a facies association consisting of volcanic-clast conglomerate, coarse arkosic sandstone, and dark red clay-rich siltstone is inferred between the sections at Hungarian Falls (along Dover Creek and a tributary falls), Hammel Creek, and Natural Wall \cite{Brojanigo1984a}. The stratigraphic correlation between Jacobsville sections at Agate Falls and Snake Creek tributary with these correlated sections from the central Keweenaw Peninsula is relatively uncertain.}
\label{fig:strat_column}
\end{figure*}

In southern exposures of the Jacobsville Formation, conglomerate facies typically have provenance sourced from Paleoproterozoic lithologies such as vein quartz and iron formation clasts \cite{Hamblin1958a, Kalliokoski1982a}. The occurrence of these conglomerates in proximity to the Marenisco fault and Pelton Creek fault is interpreted to represent syn-depositional development of local relief (Fig. \ref{fig:Geologic_map}; \citeA{Kalliokoski1982a, Hedgman2014a}). Close to the Keweenaw fault in the central Keweenaw Peninsula (Fig. \ref{fig:Geologic_map}), the Jacobsville Formation consists dominantly of quartz-rich, trough cross-bedded, medium-grained sandstone, with locally abundant clast-supported pebble to cobble conglomerate, and interbeds of red, hematite-bearing, micaceous siltstone to fine-grained sandstone (Hungarian Falls, Hammel Creek, Natural Wall; Figs. \ref{fig:strat_column} and \ref{fig:Field_photo}). In this region, the conglomerate contains abundant rounded to angular volcanic clasts as large as boulders that are likely derived from uplifted Midcontinent Rift volcanics \cite{Irving1885a, Brojanigo1984a}. Thicker intervals of dark red to brick red siltstone and very fine-grained sandstone are interbeded with coarser-grained sandstone and conglomerate at Agate Falls and Hungarian Falls (Figs. \ref{fig:strat_column} and \ref{fig:Field_photo}). Rip-up clasts of the siltstones are found within some channelized interbeds of sandstone and conglomerate (Fig. \ref{fig:intraclast_pmag}).

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{Field_photo.png}
\caption{\scriptsize Field photos of the Jacobsville Formation. (A) The red/white medium-grained sandstone with decimeter-scale trough cross-stratification in this image is a wide-spread lithology in the formation, but was not targeted for paleomagnetic sampling in this study. (B, C, D) At the Snake Creek tributary and the Natural Wall ravine in the northeastern Keweenaw Peninsula, intervals of red fine-grained sandstone to siltstone beds can be found through the steeply dipping to overturned basal strata near the Keweenaw fault and also the nearly horizontal upper strata. At the Hungarian Falls section (E), relatively thick red siltstone horizons interbedded or interfingering with sandstones and conglomerates are found along a waterfall associated with a stream that is a tributary to Dover Creek. Abundant reoriented siltstone rip-up clasts can be present within the conglomeratic layers above siltstone beds (Fig. \ref{fig:intraclast_pmag}). (F) At Agate Falls, nearly flat-lying red fissile siltstone beds are exposed. Subrounded detrital mica grains deposited parallel to the bedding plane are often present within the siltstone facies. Another set of siltstone rip-up clasts were sampled from a conglomeratic layer at Agate Falls for a paleomagnetic conglomerate test (Fig. \ref{fig:intraclast_pmag}).}
\label{fig:Field_photo}
\end{figure*}

The abundance of rift volcanic-clasts in the conglomeratic facies at Hungarian Falls, Hammel Creek, and Natural Wall indicates the presence of uplifted volcanics in proximity to the Keweenaw fault in the central part of the peninsula during Jacobsville deposition (Figs. \ref{fig:Geologic_map} and \ref{fig:strat_column}). Coarse-grained arkosic to lithic sandstone is a facies associated with the volcanic-clast conglomerates and likely derived from mechanical weathering of the exhumed volcanics. Chemical weathering of exhumed volcanics from the hanging wall is a likely source of clay within the abundant fine-grained lithologies that are associated with the volcanic-clast conglomeratic facies \cite{Hamblin1958a}. Detrital mica grains up to $\sim$5 mm in size commonly found within siltstone to fine-grained sandstone are likely sourced from Paleoproterozoic and Archean lithologies in the region (Fig. \ref{fig:Geologic_map}).

The thickness of the Jacobsville strata can vary significantly in northern Michigan \cite{Hamblin1958a, Kalliokoski1982a}, and complete stratigraphic sections are not exposed at the studied localities (Figs. \ref{fig:Geologic_map} and \ref{fig:strat_column}). Clay-rich red siltstone is more abundant in conjunction with conglomerates (e.g. 20-120 m in the Hungarian Falls section; Figs. \ref{fig:strat_column} and \ref{fig:Field_photo}) than with the cross-stratified sandstone which dominates much of the formation. Relatively thin lithic to arkosic sandstone beds, which are otherwise uncommon in the Jacobsville Formation, are commonly associated with the volcanic conglomerates and red siltstones near the Keweenaw Fault (Fig. \ref{fig:strat_column}). Together, the volcanic conglomerates, red siltstones, and arkosic sandstones make up a unique facies association. As previously interpreted by \citeA{Brojanigo1984a}, this facies association was likely deposited proximal to actively uplifting volcanics near the Hungarian Falls, Hammel Creek, and Natural Wall sections (Fig. \ref{fig:strat_column}). The relative stratigraphic positions of the Agate Falls and Snake Creek tributary sections are less certain (Figs. \ref{fig:Geologic_map} and \ref{fig:strat_column}).
